Exploring SSA Optimization Techniques

### 2014-11-12 doc/paper.tex
Added appendix A

\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, listingsutf8, mathtools}

\usepackage{hyperref}
\usepackage{breakurl}

@@

\newpage
\appendix
\section{Implementation Details for acc}

>>>>

\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, listingsutf8, mathtools}
\usepackage{hyperref}
\usepackage{breakurl}

@@

\newpage
\appendix
\section{Implementation Details for acc}
\subsection{Introduction}
  \textit{acc} (the antonijn/Antonie C Compiler) is a software project with the intent of
one-day being self-hosting (able to compile itself). The only external library
it depends on is the C99 standard library, and it's written in portable standard C99.

  acc implements many of the optimizations mentioned in the main paper, and could
serve as a reference implementation for them. It is however, much more than that,
of course, since it has to provide not only an optimizer, but back-ends and a C
front-end as well. Only the subsystems relevant to compiler optimization are
described here in detail. The most relevant subsystem is the so-called
\textit{intermediate} subsystem (shortened to \verb+itm+ in code), implementing
functions and data structures for defining and manipulating an intermediate form
SSA tree. It also implements functions for writing such a tree to a file in text
form.

  \subsection{Object oriented programming}
Although the C language doesn't natively feature object oriented syntax, it doesn't
exclude the possibility of writing clean object oriented code. For instance, the
following hierarchy:

INSERT DIAGRAM

Could well be implemented in C as follows:

\begin{lstlisting}
struct A {
	/* pointer to a B or C object */
	void *extended;

	void (*free)(struct A *self);
};

struct B {
	struct A base;

	int field;
};

struct C {
	struct B base;

	float field;
};
\end{lstlisting}

This style will be found a lot in the acc source code, although sometimes
missing the \verb+extended+ field in a base class (in which case the addresses
of both types are presumed compatible).

  \subsection{The AST and its elements}
The \verb+itm+ abstract syntax tree (AST) is not very complex. It mostly uses
linked lists (the \verb+struct list *+, for instance, is a linked list containing
only \verb+void *+ instances) for chaining instructions and blocks together.

The following intermediate code is internally represented through a few different
data structures:

\begin{lstlisting}
global int (int, int)* @gcd(int p0, int p1) {
L0:	jmp(L1);
L1:	int t2 = phi(L0, p0, L5, t7);
	int t3 = phi(L0, p1, L8, t9);
	_Bool t4 = cmp neq(t2, t3);
	split(t4, L6, L10);
L5:	_Bool t6 = cmp gt(t2, t3);
	int t7 = sub(t2, t3);
	split(t6, L1, L8);
L8:	int t9 = sub(t3, t2);
	jmp(L1);
L10:	ret(t2);
}
\end{lstlisting}

Global variables (as of yet unimplemented) and functions are represented through
\textit{container} structures (\verb+struct itm_container *+). They contain an
entry block, and are expressions themselves (as required to be called):

\begin{lstlisting}
struct itm_container {
	/* expression base */
	struct itm_expr base;

	/* container identifier */
	char *id;
	/* entry block */
	struct itm_block *block;
};
\end{lstlisting}

Blocks are represented through \verb+struct itm_block *+ structures. They are
expressions (as required to be a parameter to \verb+jmp()+ or \verb+split()+),
contain a pointer to the first instruction, the last instruction (terminal
instruction), a pointer to the block that's lexically next (\verb+L1+ for
\verb+L0+ in the first example), a pointer to the block that's lexically previous
(\verb+L0+ for \verb+L1+ in the first example, \verb+NULL+ for \verb+L0+),
and two lists of blocks that are sementically next and previous (\verb+L1+ is
\verb+L10+'s semantic predecessor, for example).

\begin{lstlisting}
struct itm_block {
	/* expression base */
	struct itm_expr base;

	/* the container the block's contained by */
	struct itm_container *container;
	/* first and last instructions */
	struct itm_instr *first, *last;
	/* blocks lexically next and previous */
	struct itm_block *lexnext, *lexprev;
	/* predecessor and successor lists */
	struct list *next, *prev;
};
\end{lstlisting}
